Structural analysis

### 2024-11-06 workflow/notebooks/draft.Rmd
BIG rework: now it can accept a list of ids or an homolog set

```{r libraries, message=FALSE, warning=FALSE, cache=FALSE}
library(tidyverse)
library(patchwork)
library(ggtree)
library(cowplot)
library(glue)
source("workflow/scripts/functions.R")
theme_set(theme_classic())

@@

params <- yaml::read_yaml("config/params.yaml")
params_dataset <- yaml::read_yaml("config/Hsapopi_test.yaml")

outdir <- glue(params$outdir, params_dataset$dataset)
```

```{r read_stuff, cache=FALSE}
seeds <- readLines(glue(outdir, "/ids/UP000005640_common.ids"))

tax <- read_delim(glue(outdir, "/meta/", params_dataset$dataset, "_uniprot_genomes.tsv"), 
                  show_col_types=F)
taxmap <- read_delim(glue(outdir, "/db/taxidmap"), 
                     delim = "\t", col_names = c("label", "Tax_ID"),
                     show_col_types = FALSE) %>% 
  left_join(tax, by = "Tax_ID") %>% 
  dplyr::select(label, Proteome_ID, mnemo)

taxidmap <- read_delim(glue(outdir,"/db/taxidmap"), 
                       col_names = c("target", "Tax_ID"), show_col_types=F)

@@

# read input table
# lvls_tax <- rev(c(unique(tax$Clade), "archaea", "bacteria"))

@@

# table <- read_delim(params_dataset$input_table_file, show_col_types=F, 
#                     delim = "\t", col_names = table_columns) %>% 
#   left_join(tax, by = "Tax_ID") %>% 
#   mutate(Clade = ifelse(is.na(Clade), mnemo, Clade),
#          Clade = factor(Clade, levels = lvls_tax))

@@

aln_seeds <- readLines(glue(outdir,"/ids/", params_dataset$seed, "_aln.ids"))
```

@@

```{r read_self_blast}
blast_self <- read_delim(glue(outdir, "/homology/allvall/",
                              params_dataset$seed, "_", params_dataset$seed, "_blast.tsv"),
                         show_col_types=FALSE,
                         col_names = blast_columns) %>% 
  # filter(query %in% seeds) %>%
  filter(query==target) %>% 
  mutate(blast_selfbit=bitscore) %>% 
  select(query, blast_selfbit)

fs_self <- read_delim(glue(outdir, "/homology/allvall/",
                              params_dataset$seed, "_", params_dataset$seed, "_fs.tsv"), 
                      show_col_types = FALSE,
                         col_names = blast_columns) %>% 
  # filter(query %in% seeds) %>%
  filter(query==target) %>% 
  mutate(fs_selfbit=bitscore) %>% 
  select(query, fs_selfbit)

@@

```{r read_fs}
fs <- read_delim(glue(outdir, "/homology/", params_dataset$seed, "_fs.tsv"), 
                 col_names = fs_columns, show_col_types = FALSE) %>% 
  # filter(query %in% seeds) %>%
  mutate(evalue = ifelse(evalue==0, 1e-180, evalue)) %>%
  left_join(taxidmap, by = "target")

fs_brhs <- read_delim(glue(outdir, "/homology/", params_dataset$seed, "_fs_brh.tsv"),
                     col_names = c("query", "target"), show_col_types = FALSE) %>% 
  # filter(query %in% seeds) %>%
  mutate(method="fs_brh") 
```

```{r read_blast}
blast <- read_delim(glue(outdir, "/homology/", params_dataset$seed, "_blast.tsv"), 
                    col_names = blast_columns, show_col_types = FALSE) %>% 
  # filter(query %in% seeds) %>%
  mutate(evalue = ifelse(evalue==0, 1e-180, evalue))

blast_brhs <- read_delim(glue(outdir, "/homology/", params_dataset$seed, "_blast_brh.tsv"),
                     col_names = c("query", "target"), show_col_types = FALSE) %>% 
  # filter(query %in% seeds) %>%
  mutate(method="blast_brh")
```

@@

brhs <- rbind(blast_brhs, fs_brhs) %>% 
    group_by(query, target, method) %>%
    mutate(n=1) %>%
    pivot_wider(names_from = method, values_fill = 0, values_from = n) %>%
    mutate(common = ifelse(fs_brh+blast_brh==2, 1, 0))

blast_filt <- filter(blast, 
                         # query %in% paste0("AF-", aln_seeds, "-F1"),
                         length/qlen*100>params$coverage, 
                         length/slen*100>params$coverage,
                         evalue<as.numeric(params$eval_both)) %>% 
  group_by(query) %>% 
  slice_head(n = params$max_seqs) %>% 
  ungroup()

fs_filt <- filter(fs, 
                      # query %in% paste0("AF-", aln_seeds, "-F1"),
                      qcov*100>params$coverage, 
                      tcov*100>params$coverage, 
                      evalue<as.numeric(params$eval_both)) %>% 
  group_by(query) %>% 
  slice_head(n = params$max_seqs) %>% 
  ungroup()

df_filt <- full_join(blast_filt, fs_filt,
                by = c("query", "target"), suffix=c("_Blast", "_Foldseek")) %>%
    filter(query!=target) %>% 
  left_join(self, by="query") %>%
  mutate(pident_Foldseek = 100*pident_Foldseek, 
         qcov_Foldseek = 100*qcov_Foldseek,
         singleton = case_when(is.na(evalue_Foldseek) ~ "only Blast",
                               is.na(evalue_Blast) ~ "only Foldseek", 
                               .default = "Common"))
# grouped_brhs <- brhs %>%
#   group_by(query) %>%
#   summarise(fs_brh = sum(fs_brh), blast_brh=sum(blast_brh), common_brh=sum(common))
blast_filtered <- filter(blast, 
                         query %in% paste0("AF-", aln_seeds, "-F1"),
                         length/qlen*100>params$coverage, 
                         length/slen*100>params$coverage,
                         evalue<as.numeric(params$eval_both)) %>% 
  group_by(query) %>% 
  slice_head(n = params$max_seqs) %>% 
  ungroup()

fs_filtered <- filter(fs, 
                      query %in% paste0("AF-", aln_seeds, "-F1"),
                      qcov*100>params$coverage, 
                      tcov*100>params$coverage, 
                      evalue<as.numeric(params$eval_both)) %>% 
  group_by(query) %>% 
  slice_head(n = params$max_seqs) %>% 
  ungroup()

df_filtered <- full_join(blast_filtered, fs_filtered,
                by = c("query", "target"), suffix=c("_Blast", "_Foldseek")) %>%
    filter(query!=target) %>% 
  left_join(self, by="query") %>%
  mutate(pident_Foldseek = 100*pident_Foldseek, 
         qcov_Foldseek = 100*qcov_Foldseek,
         singleton = case_when(is.na(evalue_Foldseek) ~ "only Blast",
                               is.na(evalue_Blast) ~ "only Foldseek", 
                               .default = "Common"))
```

```{r cath}
CATH_df <- read_delim(list.files(glue(params$data_dir, "cath"), full.names = TRUE),
                      delim = "\t", id = "proteome", show_col_types = FALSE) %>%
  # filter(!is.na(Gene3D)) %>% 
  mutate(proteome = gsub("_cath.tsv", "", basename(proteome)),
         n_3d_domains = str_count(Gene3D, ";"),
         n_Pfam_domains = str_count(Pfam, ";")) %>%
  # filter(n_domains==1) %>% 
  mutate(target=paste0("AF-", Entry, "-F1")) %>% 
    select(-proteome, Entry)

nms <- read_delim("resources/cath-names.txt", 
                  comment = "#", delim = "    ",
                  col_names = c("Gene3D", "abbr", "name"), 
                  show_col_types = FALSE) %>% 
  select(-abbr) %>% 
  mutate(name=gsub("^:", "", name))

nms_pfam <- read_delim("resources/Pfam-A.clans.tsv", 
           col_names = c("Pfam", "clan", "boh", "abbr", "name"), 
           show_col_types = FALSE) %>% 
  select(-name, -clan, -boh)

long_CATH_df <- CATH_df %>% 
  # filter(target %in% unique(blast$query)) %>% 
  mutate(Gene3D=strsplit(Gene3D, ";")) %>% 
  unnest(Gene3D) %>% 
  separate(col = Gene3D, into = c("C", "A", "T", "H"), sep = "\\.", remove = FALSE) %>% 
  mutate(CA = paste(C,A,sep = "."), CAT=paste(C,A,T,sep = ".")) %>% 
  left_join(nms, by=c("C"="Gene3D")) %>%  
  left_join(nms, by=c("CA"="Gene3D"), suffix = c("_C", "_CA")) %>% 
  left_join(nms, by=c("CAT"="Gene3D"), suffix = c("_CA", "_CAT")) %>% 
  left_join(nms, by=c("Gene3D"), suffix = c("_CAT", "_CATH"))

long_CATH_nms <- long_CATH_df %>%
    select(-c(Entry, Pfam, n_Pfam_domains, target, n_3d_domains)) %>% 
    distinct()
```

@@

```{r fig1}
fig1a <- df %>%
  group_by(singleton) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  mutate(prop=n/sum(n), Filtering="Before") %>%
  bind_rows(df_filt %>%
            group_by(singleton) %>%
            summarise(n = n()) %>%
            ungroup() %>%
            mutate(prop=n/sum(n), Filtering="After")) %>%
  mutate(Filtering=factor(Filtering, labels = c("After", "Before"), ordered = T),
         singleton=case_when(singleton=="only Blast"~"Only Bp", 
                             singleton=="only Foldseek" ~ "Only Fs",
                             .default = singleton)) %>%
  ggplot(aes(singleton, n, fill=singleton)) +
  geom_bar(aes(alpha=Filtering), stat = "identity", position = position_dodge2(reverse=TRUE)) +
  geom_text(aes(group=Filtering, label=paste0(round(prop, 2)*100, "%")),
            position = position_dodge2(reverse = TRUE, width = 0.9),
            vjust=1.3, color="white", size=3) +
  scale_fill_manual(values = palette_singleton_s, guide="none") +
  scale_alpha_manual(values = c(1,.6)) +
  scale_y_continuous(expand = c(0,0), labels = scales::label_number(scale = 1e-6)) +
  labs(fill="Class", x="Class", y="# aligned protein pairs (Mln)") +
  theme(legend.position = "bottom", legend.box="vertical")

@@

ggsave("draft/figures/fig1_lddt.pdf", fig1b, width = 3, height = 4)
ggsave("draft/figures/fig1_cnts.pdf", fig1a, width = 4, height = 2)
```

@@

df_multi <- df_filt %>% 
  select(query, target, singleton) %>% 
    head(n = 1000) %>%
  left_join(CATH_df, by=c("query"="target")) %>% 
  filter(!is.na(Pfam), !is.na(Gene3D)) %>% 
  left_join(CATH_df, by="target") %>% 
  rowwise() %>% 
  mutate(Gene3D.x=strsplit(Gene3D.x, ";"),
         Pfam.x=strsplit(Pfam.x, ";"),
         Gene3D.y=strsplit(Gene3D.y, ";"),
         Pfam.y=strsplit(Pfam.y, ";")) %>% 
  mutate(n_3d_common=length(intersect(Gene3D.x,Gene3D.y)),
         n_Pfam_common=length(intersect(Pfam.x,Pfam.y))) %>% 
  # filter(n_Pfam_common>0, n_3d_common>0) %>% 
  select(-c(Pfam.x,Pfam.y,Gene3D.x,Gene3D.y)) %>% 
  mutate(Jaccard_3d=n_3d_common/pmax(n_3d_domains.x,n_3d_domains.y),
         Jaccard_Pfam=n_Pfam_common/pmax(n_Pfam_domains.x,n_Pfam_domains.y)) %>% 
  select(query, target, singleton, contains("Jaccard"), contains("common"))

>>>>

```{r libraries, message=FALSE, warning=FALSE, cache=FALSE}
library(tidyverse)
library(patchwork)
library(ggtree)
library(cowplot)
library(glue)
source("workflow/scripts/functions.R")
theme_set(theme_classic(base_family = "Aptos"))

@@

params <- yaml::read_yaml("config/params.yaml")
params_dataset <- yaml::read_yaml("config/Hsap_draft.yaml")

homodir <- glue(params$outdir, "homology/", params_dataset$homology_dataset)
phylodir <- glue(params$outdir, "phylogeny/", params_dataset$phylo_dataset)
```

```{r read_stuff, cache=FALSE}
seeds <- readLines(params_dataset$test_seeds)

tax <- read_delim(glue("results/data/meta/",
                       params_dataset$homology_dataset, "_uniprot_genomes.tsv"), 
                  show_col_types=F)
taxmap <- read_delim(glue(homodir, "/db/taxidmap"), 
                     delim = "\t", col_names = c("label", "Tax_ID"),
                     show_col_types = FALSE) %>% 
  left_join(tax, by = "Tax_ID") %>% 
  dplyr::select(label, Proteome_ID, mnemo)

taxidmap <- read_delim(glue(homodir,"/db/taxidmap"), 
                       col_names = c("target", "Tax_ID"), show_col_types=F)

@@

@@

@@

aln_seeds <- readLines(glue(phylodir,"/ids/", params_dataset$seed, "_aln.ids"))
```

@@

```{r read_self_blast}
blast_self <- read_delim(glue(homodir, "/allvall/",
                              params_dataset$seed, "_", params_dataset$seed, "_blast.tsv"),
                         show_col_types=FALSE,
                         col_names = blast_columns) %>% 
  # filter(query %in% seeds) %>%
  filter(query==target) %>% 
  mutate(blast_selfbit=bitscore) %>% 
  select(query, blast_selfbit)

fs_self <- read_delim(glue(homodir, "/allvall/", params_dataset$seed, 
                           "_", params_dataset$seed, "_fs.tsv"), 
                      show_col_types = FALSE,
                         col_names = blast_columns) %>% 
  # filter(query %in% seeds) %>%
  filter(query==target) %>% 
  mutate(fs_selfbit=bitscore) %>% 
  select(query, fs_selfbit)

@@

```{r read_homo}
fs <- read_delim(glue(homodir, "/", params_dataset$seed, "_fs.tsv"), 
                 col_names = fs_columns, show_col_types = FALSE) %>% 
  # filter(query %in% seeds) %>%
  mutate(evalue = ifelse(evalue==0, 1e-180, evalue)) %>%
  left_join(taxidmap, by = "target")

blast <- read_delim(glue(homodir, "/", params_dataset$seed, "_blast.tsv"), 
                    col_names = blast_columns, show_col_types = FALSE) %>% 
  # filter(query %in% seeds) %>%
  mutate(evalue = ifelse(evalue==0, 1e-180, evalue))
```

@@

blast_filt <- filter(blast, 
                         length/qlen*100>params$coverage, 
                         length/slen*100>params$coverage,
                         evalue<as.numeric(params$eval_both)) %>% 
  group_by(query) %>% 
  slice_head(n = params$max_seqs) %>% 
  ungroup()

fs_filt <- filter(fs, 
                      qcov*100>params$coverage, 
                      tcov*100>params$coverage, 
                      evalue<as.numeric(params$eval_both)) %>% 
  group_by(query) %>% 
  slice_head(n = params$max_seqs) %>% 
  ungroup()

df_filt <- full_join(blast_filt, fs_filt,
                by = c("query", "target"), suffix=c("_Blast", "_Foldseek")) %>%
    filter(query!=target) %>% 
  left_join(self, by="query") %>%
  mutate(pident_Foldseek = 100*pident_Foldseek, 
         qcov_Foldseek = 100*qcov_Foldseek,
         singleton = case_when(is.na(evalue_Foldseek) ~ "only Blast",
                               is.na(evalue_Blast) ~ "only Foldseek", 
                               .default = "Common"))

# blast_filtered <- filter(blast, 
#                          query %in% paste0("AF-", aln_seeds, "-F1"),
#                          length/qlen*100>params$coverage, 
#                          length/slen*100>params$coverage,
#                          evalue<as.numeric(params$eval_both)) %>% 
#   group_by(query) %>% 
#   slice_head(n = params$max_seqs) %>% 
#   ungroup()
# 
# fs_filtered <- filter(fs, 
#                       query %in% paste0("AF-", aln_seeds, "-F1"),
#                       qcov*100>params$coverage, 
#                       tcov*100>params$coverage, 
#                       evalue<as.numeric(params$eval_both)) %>% 
#   group_by(query) %>% 
#   slice_head(n = params$max_seqs) %>% 
#   ungroup()
# 
# df_filtered <- full_join(blast_filtered, fs_filtered,
#                 by = c("query", "target"), suffix=c("_Blast", "_Foldseek")) %>%
#     filter(query!=target) %>% 
#   left_join(self, by="query") %>%
#   mutate(pident_Foldseek = 100*pident_Foldseek, 
#          qcov_Foldseek = 100*qcov_Foldseek,
#          singleton = case_when(is.na(evalue_Foldseek) ~ "only Blast",
#                                is.na(evalue_Blast) ~ "only Foldseek", 
#                                .default = "Common"))
```

```{r cath}
CATH_df <- read_delim(paste0(params$data_dir, "cath/", tax$Proteome_ID, "_cath.tsv"),
                      delim = "\t", id = "proteome", show_col_types = FALSE) %>%
  mutate(proteome = gsub("_cath.tsv", "", basename(proteome))) %>%
         # n_3d_domains = str_count(Gene3D, ";"),
         # n_Pfam_domains = str_count(Pfam, ";")
  mutate(target=paste0("AF-", Entry, "-F1")) %>% 
  rename("CATH"="Gene3D") %>% 
  select(target, Pfam, CATH, -proteome, -Entry) %>% 
  mutate(CAT = gsub("(\\d+\\.\\d+\\.\\d+)\\.\\d+", "\\1", CATH),
         CA = gsub("(\\d+\\.\\d+)\\.\\d+", "\\1", CAT),
         C = gsub("\\.\\d+", "\\1", CA))

ortho_dbs <- read_delim(paste0(params$data_dir, "ids/", tax$Proteome_ID, "_ids.tsv"),
                      delim = "\t", id = "proteome", show_col_types = FALSE) %>% 
  mutate(proteome = gsub("_ids.tsv", "", basename(proteome))) %>% 
  mutate(target=paste0("AF-", Entry, "-F1")) %>% 
  select(-proteome, -Entry)

ortho_dbs <- full_join(CATH_df, ortho_dbs)

# nms <- read_delim("resources/cath-names.txt", 
#                   comment = "#", delim = "    ",
#                   col_names = c("Gene3D", "abbr", "name"), 
#                   show_col_types = FALSE) %>% 
#   select(-abbr) %>% 
#   mutate(name=gsub("^:", "", name))
# 
# nms_pfam <- read_delim("resources/Pfam-A.clans.tsv", 
#            col_names = c("Pfam", "clan", "boh", "abbr", "name"), 
#            show_col_types = FALSE) %>% 
#   select(-name, -clan, -boh)
# 
# long_CATH_df <- CATH_df %>% 
#   # filter(target %in% unique(blast$query)) %>% 
#   mutate(Gene3D=strsplit(Gene3D, ";")) %>% 
#   unnest(Gene3D) %>% 
#   separate(col = Gene3D, into = c("C", "A", "T", "H"), sep = "\\.", remove = FALSE) %>% 
#   mutate(CA = paste(C,A,sep = "."), CAT=paste(C,A,T,sep = ".")) %>% 
#   left_join(nms, by=c("C"="Gene3D")) %>%  
#   left_join(nms, by=c("CA"="Gene3D"), suffix = c("_C", "_CA")) %>% 
#   left_join(nms, by=c("CAT"="Gene3D"), suffix = c("_CA", "_CAT")) %>% 
#   left_join(nms, by=c("Gene3D"), suffix = c("_CAT", "_CATH"))
# 
# long_CATH_nms <- long_CATH_df %>%
#     select(-c(Entry, Pfam, n_Pfam_domains, target, n_3d_domains)) %>% 
#     distinct()
```

@@

```{r fig1}
fig1a <- df %>%
  group_by(singleton) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  mutate(prop=n/sum(n), Filtering="Before") %>%
  bind_rows(df_filt %>%
            group_by(singleton) %>%
            summarise(n = n()) %>%
            ungroup() %>%
            mutate(prop=n/sum(n), Filtering="After")) %>%
  mutate(Filtering=factor(Filtering, labels = c("After", "Before"), ordered = T),
         singleton=case_when(singleton=="only Blast"~"Only Bp", 
                             singleton=="only Foldseek" ~ "Only Fs",
                             .default = singleton)) %>%
  ggplot(aes(singleton, n, fill=singleton)) +
  geom_bar(aes(alpha=Filtering), stat = "identity", position = position_dodge2(reverse=TRUE)) +
  # geom_text(aes(group=Filtering, label=paste0(round(prop, 2)*100, "%")),
  #           position = position_dodge2(reverse = TRUE, width = 0.9),
  #           vjust=1.3, color="white", size=3) +
  scale_fill_manual(values = palette_singleton_s, guide="none") +
  scale_alpha_manual(values = c(1,.6)) +
  scale_y_continuous(expand = c(0,0), labels = scales::label_number(scale = 1e-6)) +
  labs(fill="Class", x="Class", y="Milions of aligned protein pairs") +
  theme(legend.position = "bottom", legend.box="vertical")

@@

# ggsave("draft/figures/fig1_lddt.pdf", fig1b, width = 3, height = 4)
# ggsave("draft/figures/fig1_cnts.pdf", fig1a, width = 4, height = 2)
```

@@

dbs <- colnames(ortho_dbs)[2:dim(ortho_dbs)[2]]
# Loop through each column name and apply a function
jaccard_df <- map_dfr(dbs, ~ {
  col_name <- .x
select(df_filt, query, target, singleton) %>% 
    # filter(query=="AF-A0A024RBG1-F1") %>%
    left_join(select(ortho_dbs, target, col_name), by=c("query"="target")) %>% 
    left_join(select(ortho_dbs, target, col_name), by=c("target")) %>% 
    rename("query_set"=paste0(col_name, ".x"), "target_set"=paste0(col_name, ".y")) %>% 
    # Exclude queries without any domain
    filter(!is.na(query_set), !is.na(target_set)) %>% 
    group_by(query, singleton) %>%
    distinct(query, singleton, query_set, target_set) %>% 
    summarise(set_q = strsplit(paste(query_set, collapse = ""), ";"),
              set_t = strsplit(paste(target_set, collapse = ""), ";")) %>%
    rowwise() %>%
    mutate(jaccard = length(intersect(set_q, set_t))/length(union(set_q, set_t))) %>% 
    select(-set_q, -set_t) %>% 
    mutate(db=col_name)

})

ggplot(jaccard_df, aes(jaccard, fill=singleton)) +
  geom_histogram(position = "dodge") +
  facet_wrap(~db, scales = "free_y") +
  scale_y_continuous(expand = expansion(0.01,0)) + 
  scale_fill_manual(values=palette_singleton_p) +
  theme_bw() +
  theme(legend.position = "top",
        # panel.background = element_rect(fill=NA, color="black"),
        # axis.line = element_blank()
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), 
        strip.background = element_rect(fill=NA)
        # panel.grid.major.x = element_line()
        )

ggsave("test/bench.pdf", width = 8, height = 5, device = cairo_pdf)
# plot_dbs_list[[db]] <- p_db

# df_multi <- df_filt %>% 
#   select(query, target, singleton) %>% 
#     head(n = 100) %>%
#   left_join(CATH_df, by=c("query"="target")) %>% 
#   filter(!is.na(Pfam), !is.na(Gene3D)) %>% 
#   left_join(CATH_df, by="target") %>% 
#   rowwise() %>% 
#   mutate(Gene3D.x=strsplit(Gene3D.x, ";"),
#          Pfam.x=strsplit(Pfam.x, ";"),
#          Gene3D.y=strsplit(Gene3D.y, ";"),
#          Pfam.y=strsplit(Pfam.y, ";")) %>% 
#   mutate(n_3d_common=length(intersect(Gene3D.x,Gene3D.y)),
#          n_Pfam_common=length(intersect(Pfam.x,Pfam.y))) %>% 
#   # filter(n_Pfam_common>0, n_3d_common>0) %>% 
#   select(-c(Pfam.x,Pfam.y,Gene3D.x,Gene3D.y)) %>% 
#   mutate(Jaccard_3d=n_3d_common/pmax(n_3d_domains.x,n_3d_domains.y),
#          Jaccard_Pfam=n_Pfam_common/pmax(n_Pfam_domains.x,n_Pfam_domains.y)) %>% 
#   select(query, target, singleton, contains("Jaccard"), contains("common"))
